Depletion capabilities in the OpenMC Monte Carlo particle transport code

### 2020-07-10 manuscript.tex
Incorporate feedback on section 2 from @cjosey

The equation that governs the transmutation of nuclides by nuclear reactions and
radioactive decay inside of an irradiated environment can be written as
\begin{equation}
  \label{eq:depletion}
  \begin{split}
  \frac{dN_i(t)}{dt} = &\sum\limits_j \left [ f_{j \rightarrow i}
  \int_0^\infty dE \; \sigma_j (E, t) \phi(E,t) + \lambda_{j\rightarrow i}
  \right ] N_j(t) \\ &- \left [\int_0^\infty dE \; \sigma_i (E,t) \phi(E,t) +
  \sum\limits_j \lambda_{i\rightarrow j} \right ] N_i(t), \quad i=1,\dots,n
  \end{split}
\end{equation}
where
\begin{equation*}
  \begin{split}
  N_i(t) &\equiv \text{density of nuclide $i$ at time $t$} \\
  \sigma_i(E,t) &\equiv \text{transmutation cross section for nuclide $i$ at energy $E$ and time $t$} \\
  \phi(E,t) &\equiv \text{neutron flux at energy $E$ and time $t$} \\
  f_{j \rightarrow i} &\equiv \text{fraction of transmutation reactions in nuclide $j$ that produce nuclide $i$} \\
  \lambda_{j \rightarrow i} &\equiv \text{decay constant for decay modes in nuclide $j$ that produce nuclide $i$} \\
  n &\equiv \text{total number of nuclides.}
  \end{split}
\end{equation*}
Note that we have not included the spatial dependence of the flux or cross
sections. For \cref{eq:depletion} to be accurate, the spatial variation of the
reaction rates over the region which is being depleted should be small. For a
reactor model, this may entail using a unique material in each fuel pin that is
depleted separately.\footnote{OpenMC includes functions for automatically
differentiating materials in cells that are repeated throughout a geometry.}
Ultimately, the user is responsible for deciding what level of spatial
discretization is adequate for the problem at hand.

@@

Although the neutron transport equation is time-dependent, the timescale over
which material compositions change is sufficiently long such that the transport
equation can be solved as a steady state equation. Without loss of generality,
let us write the burnup matrix solely as a function of $\vect{n}$ so that
\cref{eq:depletion-matrix-t} becomes
\begin{equation}
  \label{eq:depletion-matrix}
  \frac{d\vect{n}}{dt} = \vect{A}(\vect{n})\vect{n}, \quad \vect{n}(0) =
  \vect{n}_0.
\end{equation}
This better reflects the fact that the time, $t$, is generally not an input to
the transport equation, and hence the burnup matrix is uniquely determined by
the value of $\vect{n}$.

@@

OpenMC doesn't rely on a single time integration method but instead has several
methods available that the user can choose. In the following sections, we
outline the available methods in OpenMC, several of which have not been
previously described in the literature.

@@

The Magnus expansion can be used as the basis for developing numerical
integration algorithms and has been widely studied. The typical approach is to
truncate the $\vect{\Omega}$ series at an appropriate order and apply an
approximation to the multivariate integrals in \cref{eq:magnus-terms}. In
particular, OpenMC uses a fourth-order commutator-free\footnote{The integrator
is called commutator-free because it does not require computing any matrix
commutators like those that appear in \cref{eq:magnus-terms}. Other methods
based on the Magnus expansion do require computing matrix commutators, some of
which were explored by \citet{josey2017phd}.} exponential
integrator~\citep{blanes2006anm} that is derived from the Magnus expansion in
order to solve \cref{eq:linear-ode}. In this approximation, the Magnus expansion
is truncated at two terms and then expressed as the composition of two
exponentials. For a system of ODEs in the form of \cref{eq:linear-ode}, the
resulting method is
\begin{equation}
  \label{eq:compose4}
  \vect{n}_{i+1} = \exp (a_2 \vect{A}_1 + a_1 \vect{A}_2)
  \exp (a_1 \vect{A}_1 + a_2 \vect{A}_2) \vect{n}_i
\end{equation}
where $\vect{A}_1$ and $\vect{A}_2$ correspond to $\hat{\vect{A}}$ evaluated at
the nodes of a two-point Gauss-Legendre quadrature over the interval $[0,h]$:
\begin{equation}
  \begin{split}
    a_1 &= \frac{1}{4} + \frac{\sqrt{3}}{6}, \; a_2 = \frac{1}{4} - \frac{\sqrt{3}}{6} \\
    c_1 &= \frac{1}{2} - \frac{\sqrt{3}}{6}, \; c_2 = \frac{1}{2} + \frac{\sqrt{3}}{6} \\
    \vect{A}_1 &= h\hat{\vect{A}}\left ( c_1 h\right ), \; \vect{A}_2 = h\hat{\vect{A}}\left ( c_2 h\right ).
  \end{split}
\end{equation}
Evaluating $\vect{A}_1$ and $\vect{A}_2$ using \cref{eq:linear-interp} and
substituting in \cref{eq:compose4}, we obtain a CE/LI method based on the
fourth-order commutator-free Magnus integrator:
\begin{equation}
  \label{eq:celi-cfq4}
  \begin{split}
  \vect{n}_{i+1}^p &= \exp \left ( h \vect{A}(\vect{n}_i ) \right ) \\
  \vect{n}_{i+1} &= \exp \left( \frac{h}{12} \vect{A}(\vect{n}_i) +
    \frac{5h}{12} \vect{A}(\vect{n}_{i+1}^p) \right)
  \exp \left( \frac{5h}{12} \vect{A}(\vect{n}_i) +
  \frac{h}{12} \vect{A}(\vect{n}_{i+1}^p) \right) \vect{n}_i.
  \end{split}
\end{equation}
\Cref{eq:celi-cfq4} requires two transport solutions and two matrix
exponentials, whereas the CE/LI method with substeps, \cref{eq:celi-substeps},
requires two transport solutions and $m$ matrix exponentials. A comparison of
these approaches has shown that \cref{eq:celi-cfq4} achieves better accuracy
than the use of five substeps~\citep{josey2017phd}. OpenMC also includes an
integrator based on the LE/QI (linear extrapolation, quadratic interpolation)
method~\citep{isotalo2011ane2} where the integration of \cref{eq:linear-ode} (in
which case $\hat{\vect{A}}(t)$ is represented by a quadratic polynomial in $t$)
is performed using the fourth-order commutator-free Magnus integrator based on
\cref{eq:compose4}.

@@

All of the integration methods discussed thus far are explicit methods that are
only conditionally stable. There are many well-documented cases where
oscillatory behavior and diverging solutions have been observed. To address
this, \citet{dufek2013ane} proposed an implicit-Euler-like method that is
unconditionally stable. The main idea is to replace $\vect{A}(\vect{n}_i)$ with
$\vect{A}(\vect{n}_{i+1})$ in \cref{eq:predictor}:
\begin{equation}
  \label{eq:implicit}
  \vect{n}_{i=1} = \exp \left( h\vect{A}(\vect{n}_{i+1}) \right) \vect{n}_i.
\end{equation}
This results in a non-linear equation for $\vect{n}_{i+1}$ that cannot be solved
directly. However, it is possible to solve for $\vect{n}_{i+1}$ using a
fixed-point iteration. \citet{dufek2013ane} utilized a stochastic approximation
based on the Robbins--Monro algorithm~\citep{robbins1951ams} in order to
guarantee convergence of the fixed-point iteration while accounting for
stochasticity in the MC transport solver.

The major downside of an algorithm based on \cref{eq:implicit} is that, like the
predictor method, it is only first-order accurate. Implicit methods have also
been applied to higher-order
algorithms~\citep{kotlyar2014ane,kotlyar2016ane,cosgrove2020ane}, thereby
obtaining methods that are unconditionally stable and accurate with longer
timesteps. In OpenMC, stochastic implicit versions of the CE/LI and LE/QI
methods have also been implemented. The basis for the algorithm is to iterate on
$\vect{n}_{i+1}^p$ used in \cref{eq:linear-interp}, obtaining progressively
better estimates that are then used to solve \cref{eq:linear-ode}. A full
discussion of the algorithm and its performance and tradeoffs can be found in
\citep{josey2017phd}.

@@

OpenMC does not rely on a single time integration method but rather has several
classes that implement different algorithms. Generally, there is a tradeoff
between the accuracy of the method, the computational expense, and the memory
requirements. When a MC code is used, the expense is dominated by the time
to compute the transport solution, i.e., to evaluate $\vect{A}$ for a given
$\vect{n}$. Thus, the cost of a method scales with the number of $\vect{A}$
evaluations that are performed per timestep. On the other hand, methods that
require more evaluations generally achieve higher accuracy. The predictor method
only requires one evaluation and its error converges as $\mathcal{O}(h)$. The
CE/CM method requires two evaluations and is thus twice as expensive as the
predictor method, but achieves an error of $\mathcal{O}(h^2)$. An exhaustive
description of time integration methods and their merits can be found in the
thesis of \citet{josey2017phd}.

@@

(n,$\gamma$) reactions may result in a product nucleus being in either the
ground state or an excited state. If the excited state is a metastable state
(having an unusually long half-life), it is important to account for this
difference during depletion. The most well-known example is capture in
$^{241}$Am, which can produce either $^{242}$Am or $^{242\text{m}}$Am. Because the
metastable state has a significantly longer half-life than the ground state, it
is important to account for the branching of the capture reaction in $^{241}$Am
to accurately predict the production of higher actinides~\citep{haeck2012nse}.
This is complicated by the fact that the branching ratio may depend on the
incident neutron energy causing capture.

OpenMC does not currently implement a method to handle energy-dependent capture
branching ratios directly. However, the depletion chain file does allow a
reaction to be listed multiple times with different branching ratios that result
in different products. Spectrum-averaged capture branching ratios have been
computed in LWR and SFR spectra and are available at
\url{https://openmc.org/depletion-chains}. Within the Python API, the class
representing a depletion chain allows branching ratios to be easily retrieved
and changed. We note that while this treatment doesn't accurately account for
the change in the capture branching ratio with increasing burnup (due to
spectral changes), these changes are generally much smaller than the differences
in capture branching ratios between nuclear data evaluations.

>>>>

The equation that governs the transmutation of nuclides by nuclear reactions and
radioactive decay inside of an irradiated environment can be written as
\begin{equation}
  \label{eq:depletion}
  \begin{split}
  \frac{dN_i(t)}{dt} = &\sum\limits_j \left [ \int_0^\infty dE \;
  f_{j \rightarrow i}(E) \sigma_j (E, t) \phi(E,t) + \lambda_{j\rightarrow i}
  \right ] N_j(t) \\ &- \left [\int_0^\infty dE \; \sigma_i (E,t) \phi(E,t) +
  \sum\limits_j \lambda_{i\rightarrow j} \right ] N_i(t), \quad i=1,\dots,n
  \end{split}
\end{equation}
where
\begin{equation*}
  \begin{split}
  N_i(t) &\equiv \text{density of nuclide $i$ at time $t$} \\
  \sigma_i(E,t) &\equiv \text{transmutation cross section for nuclide $i$ at energy $E$ and time $t$} \\
  \phi(E,t) &\equiv \text{neutron flux at energy $E$ and time $t$} \\
  f_{j \rightarrow i}(E) &\equiv \text{fraction of transmutation reactions in nuclide $j$ that produce nuclide $i$} \\
  \lambda_{j \rightarrow i} &\equiv \text{decay constant for decay modes in nuclide $j$ that produce nuclide $i$} \\
  n &\equiv \text{total number of nuclides.}
  \end{split}
\end{equation*}
Note that we have not included the spatial dependence of the flux or cross
sections. For \cref{eq:depletion} to be accurate, the spatial variation of the
reaction rates over the region which is being depleted should be small. For a
reactor model, this may entail using a unique material in each fuel pin that is
depleted separately.\footnote{OpenMC includes functions for automatically
differentiating materials in cells that are repeated throughout a geometry.}
Ultimately, the user is responsible for deciding what level of spatial
discretization is adequate for the problem at hand.

@@

Evaluating $\vect{A}$ for a given $\vect{n}$ and $t$ is equivalent to solving
the neutron transport equation to obtain transmutation reaction rates and
combining that information with decay and fission product yield data to
construct the matrix of coefficients. Although the transport equation is
time-dependent, the timescale over which material compositions change is
sufficiently long such that the transport equation can be solved as a steady
state equation. This implies that the only ``input'' that is required by the
transport solver is $\vect{n}$, as opposed to $\vect{n}$ and $t$. Without loss of
generality, let us write the burnup matrix solely as a function of $\vect{n}$ so
that \cref{eq:depletion-matrix-t} becomes
\begin{equation}
  \label{eq:depletion-matrix}
  \frac{d\vect{n}}{dt} = \vect{A}(\vect{n})\vect{n}, \quad \vect{n}(0) =
  \vect{n}_0.
\end{equation}
This better reflects the fact that the burnup matrix is uniquely determined by
the value of $\vect{n}$. To be clear, $\vect{A}$ is still a function of time but
only transitively through the dependence on $\vect{n}(t)$.

@@

OpenMC doesn't rely on a single time integration method but instead has several
methods available that the user can choose. In the following sections, we
outline the available methods in OpenMC, several of which have not been
fully described in the literature. An exhaustive analysis of these time
integration methods and their merits can be found in the thesis of
\citet{josey2017phd}.

@@

The Magnus expansion can be used as the basis for developing numerical
integration algorithms and has been widely studied. The typical approach is to
truncate the $\vect{\Omega}$ series at an appropriate order and apply an
approximation to the multivariate integrals in \cref{eq:magnus-terms}. In
particular, OpenMC uses a fourth-order commutator-free\footnote{The integrator
is called commutator-free because it does not require computing any matrix
commutators like those that appear in \cref{eq:magnus-terms}.} exponential
integrator~\citep{blanes2006anm} that is derived from the Magnus expansion in
order to solve \cref{eq:linear-ode}. In this approximation, the Magnus expansion
is truncated at two terms and then expressed as the composition of two
exponentials. For a system of ODEs in the form of \cref{eq:linear-ode}, the
resulting method is
\begin{equation}
  \label{eq:compose4}
  \vect{n}_{i+1} = \exp (a_2 \vect{A}_1 + a_1 \vect{A}_2)
  \exp (a_1 \vect{A}_1 + a_2 \vect{A}_2) \vect{n}_i
\end{equation}
where $\vect{A}_1$ and $\vect{A}_2$ correspond to $\hat{\vect{A}}$ evaluated at
the nodes of a two-point Gauss-Legendre quadrature over the interval $[0,h]$:
\begin{equation}
  \begin{split}
    a_1 &= \frac{1}{4} + \frac{\sqrt{3}}{6}, \; a_2 = \frac{1}{4} - \frac{\sqrt{3}}{6} \\
    c_1 &= \frac{1}{2} - \frac{\sqrt{3}}{6}, \; c_2 = \frac{1}{2} + \frac{\sqrt{3}}{6} \\
    \vect{A}_1 &= h\hat{\vect{A}}\left ( c_1 h\right ), \; \vect{A}_2 = h\hat{\vect{A}}\left ( c_2 h\right ).
  \end{split}
\end{equation}
Evaluating $\vect{A}_1$ and $\vect{A}_2$ using \cref{eq:linear-interp} and
substituting in \cref{eq:compose4}, we obtain a CE/LI method based on the
fourth-order commutator-free Magnus integrator:
\begin{equation}
  \label{eq:celi-cfq4}
  \begin{split}
  \vect{n}_{i+1}^p &= \exp \left ( h \vect{A}(\vect{n}_i ) \right ) \\
  \vect{n}_{i+1} &= \exp \left( \frac{h}{12} \vect{A}(\vect{n}_i) +
    \frac{5h}{12} \vect{A}(\vect{n}_{i+1}^p) \right)
  \exp \left( \frac{5h}{12} \vect{A}(\vect{n}_i) +
  \frac{h}{12} \vect{A}(\vect{n}_{i+1}^p) \right) \vect{n}_i.
  \end{split}
\end{equation}
\Cref{eq:celi-cfq4} requires two transport solutions and two matrix
exponentials, whereas the CE/LI method with substeps, \cref{eq:celi-substeps},
requires two transport solutions and $m$ matrix exponentials. A comparison of
these approaches has shown that \cref{eq:celi-cfq4} achieves better accuracy
than the use of five substeps~\citep{josey2017phd}. OpenMC also includes an
integrator based on the LE/QI (linear extrapolation, quadratic interpolation)
method~\citep{isotalo2011ane2} where the integration of \cref{eq:linear-ode} (in
which case $\hat{\vect{A}}(t)$ is represented by a quadratic polynomial in $t$)
is performed using the fourth-order commutator-free Magnus integrator based on
\cref{eq:compose4}.

@@

All of the integration methods discussed thus far are explicit methods that are
only conditionally stable. There are many well-documented cases where
oscillatory behavior and diverging solutions have been observed. To address
this, \citet{dufek2013ane} proposed an implicit-Euler-like method that is
unconditionally stable. The main idea is to replace $\vect{A}(\vect{n}_i)$ with
$\vect{A}(\vect{n}_{i+1})$ in \cref{eq:predictor}:
\begin{equation}
  \label{eq:implicit}
  \vect{n}_{i+1} = \exp \left( h\vect{A}(\vect{n}_{i+1}) \right) \vect{n}_i.
\end{equation}
This results in a non-linear equation for $\vect{n}_{i+1}$ that cannot be solved
directly. However, it is possible to solve for $\vect{n}_{i+1}$ using a
fixed-point iteration. \citet{dufek2013ane} utilized a stochastic approximation
based on the Robbins--Monro algorithm~\citep{robbins1951ams} in order to
guarantee convergence of the fixed-point iteration while accounting for
stochasticity in the MC transport solver.

The major downside of an algorithm based on \cref{eq:implicit} is that, like the
predictor method, it is only first-order accurate. Implicit methods have also
been applied to higher-order
algorithms~\citep{kotlyar2014ane,kotlyar2016ane,cosgrove2020ane}, thereby
obtaining methods that are unconditionally stable and accurate with longer
timesteps. In OpenMC, stochastic implicit versions of the CE/LI and LE/QI
methods have also been implemented. The basis for the algorithm is to iterate on
$\vect{n}_{i+1}^p$ used in \cref{eq:linear-interp}, obtaining progressively
better estimates that are then used to solve \cref{eq:linear-ode}.

@@

OpenMC does not rely on a single time integration method but rather has several
classes that implement different algorithms. Generally, there is a tradeoff
between the accuracy of the method, the computational expense, and the memory
requirements. When a MC code is used, the expense is dominated by the time
to compute the transport solution, i.e., to evaluate $\vect{A}$ for a given
$\vect{n}$. Thus, the cost of a method scales with the number of $\vect{A}$
evaluations that are performed per timestep. On the other hand, methods that
require more evaluations generally achieve higher accuracy. The predictor method
only requires one evaluation and its error converges as $\mathcal{O}(h)$. The
CE/CM method requires two evaluations and is thus twice as expensive as the
predictor method, but achieves an error of $\mathcal{O}(h^2)$.

@@

(n,$\gamma$) reactions may result in a product nucleus being in either the
ground state or an excited state. If the excited state is a metastable state
(having an unusually long half-life), it is important to account for this
difference during depletion. The most well-known example is capture in
$^{241}$Am, which can produce either $^{242}$Am or $^{242\text{m}}$Am. Because
the metastable state has a significantly longer half-life than the ground state,
it is important to account for the branching of the capture reaction in
$^{241}$Am to accurately predict the production of higher
actinides~\citep{haeck2012nse}. This is complicated by the fact that the
branching ratio may depend on the incident neutron energy causing capture (hence
why $f_{j\rightarrow i}$ in \cref{eq:depletion} depends on $E$).

OpenMC does not currently implement a method to handle energy-dependent capture
branching ratios directly and treats $f_{j\rightarrow i}$ as a constant.
However, the depletion chain file does allow a reaction to be listed multiple
times with different branching ratios that result in different products.
Spectrum-averaged capture branching ratios have been computed in LWR and SFR
spectra and are available at \url{https://openmc.org/depletion-chains}. Within
the Python API, the class representing a depletion chain allows branching ratios
to be easily retrieved and changed. We note that while this treatment doesn't
accurately account for the change in the capture branching ratio with increasing
burnup (due to spectral changes), these changes are generally much smaller than
the differences in capture branching ratios between nuclear data evaluations.
